---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = FALSE
)
```

# bricks

<!-- badges: start -->
[![R-CMD-check](https://github.com/NC-DAC-Research-and-Planning/bricks/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/NC-DAC-Research-and-Planning/bricks/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`bricks` is a small, unified interface for reading data from Databricks in R,
regardless of *how* you connect. Write the query once; switch backends with a
single option.

| backend                | how you connect                                   | compute runs        |
|------------------------|---------------------------------------------------|---------------------|
| `"odbc"`               | `DBI` + `odbc` SQL warehouse                       | locally, on pull    |
| `"databricks"`         | `sparklyr` (native, inside a notebook)             | in Databricks       |
| `"databricks_connect"` | `sparklyr` via Databricks Connect (`pysparklyr`)   | in Databricks       |
| `"csv"`                | a local governed/offline extract                   | locally             |

## Installation

```{r}
# install.packages("pak")
pak::pak("NC-DAC-Research-and-Planning/bricks")
```

## Usage

The same three lines work for every backend:

```{r}
library(bricks)

con <- bricks_connect()                        # backend auto-resolved
on.exit(bricks_disconnect(con), add = TRUE)

# Raw SQL, lazily:
dat <- bricks_sql(con, "select * from main.default.diamonds") |>
  bricks_collect()

# Or a lazy table with dplyr push-down:
dat <- bricks_tbl(con, "main.default.diamonds") |>
  dplyr::filter(cut == "Ideal") |>
  dplyr::summarise(n = dplyr::n(), .by = color) |>
  bricks_collect()
```

`bricks_tbl()` and `bricks_sql()` stay **lazy** (via `dbplyr` / `sparklyr`), so
filters, joins, and aggregations push down to Databricks compute.
`bricks_collect()` pulls a local tibble and casts Spark `BIGINT`
(`bit64::integer64`) columns to `double`.

## Choosing a backend

`bricks_connect()` resolves the backend from, in order:

1. the `bricks.backend` option — `options(bricks.backend = "odbc")`
2. the `BRICKS_BACKEND` environment variable
3. auto-detection — `"databricks"` inside a Databricks runtime, else `"odbc"`

## Configuration

Connections read credentials from the environment (e.g. `~/.Renviron`); nothing
is stored by the package. Depending on the backend you use:

* `DATABRICKS_HOST`, `DATABRICKS_WAREHOUSE_ID` — SQL warehouse (odbc)
* `DATABRICKS_HOST`, `DATABRICKS_TOKEN`, and optionally `DATABRICKS_CLUSTER_ID`
  / `DATABRICKS_DBR_VERSION` — Databricks Connect

Never commit `.Renviron` or tokens.
