---
title: "Getting started with bricks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with bricks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(bricks)
```

## The idea

`bricks` gives you one small vocabulary for reading Databricks data, so the
same analysis code runs whether you are on a SQL warehouse, inside a Databricks
notebook, connected remotely with Databricks Connect, or working from a local
CSV extract. You pick the backend once (or let it auto-detect); everything else
is identical.

## The workflow

```{r}
con <- bricks_connect()
on.exit(bricks_disconnect(con), add = TRUE)

dat <- bricks_tbl(con, "main.default.diamonds") |>
  dplyr::filter(cut == "Ideal") |>
  dplyr::summarise(mean_price = mean(price), .by = color) |>
  bricks_collect()
```

Three ideas do all the work:

* **`bricks_connect()`** opens a connection for the resolved backend and returns
  a lightweight `bricks_con` the other verbs dispatch on.
* **`bricks_tbl()` / `bricks_sql()`** return *lazy* tables. Additional `dplyr`
  verbs are translated to SQL and executed by Databricks, not by R.
* **`bricks_collect()`** materialises the result locally as a tibble, casting
  Spark `BIGINT` columns (`bit64::integer64`) to `double` on the way out.

## Choosing a backend

Resolution order, highest priority first:

```{r}
# 1. an explicit option
options(bricks.backend = "odbc")

# 2. an environment variable
Sys.setenv(BRICKS_BACKEND = "databricks_connect")

# 3. otherwise auto-detect: "databricks" on a cluster, else "odbc"
bricks_backend()
```

You can also pass the backend directly:

```{r}
con <- bricks_connect("csv", path = "extract.csv")
```

## Credentials

Connections read from the environment (for example `~/.Renviron`); the package
stores nothing itself.

* SQL warehouse (`odbc`): `DATABRICKS_HOST`, `DATABRICKS_WAREHOUSE_ID`.
* Databricks Connect: `DATABRICKS_HOST`, `DATABRICKS_TOKEN`, and optionally
  `DATABRICKS_CLUSTER_ID` / `DATABRICKS_DBR_VERSION`.

The Databricks Connect backend also needs a one-time Python environment install:

```{r}
pysparklyr::install_databricks()
```
