---
title: "Connecting from Windows (and RStudio/Positron) survival guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Connecting from Windows (and RStudio/Positron) survival guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Getting a bricks connection working from a Windows workstation trips over a few
environment issues that have nothing to do with Databricks itself. This guide
collects the ones we have actually hit, in the order they tend to bite.

Start every diagnosis with `bricks_check()`, which reports what a backend needs
without opening a connection:

```{r}
library(bricks)
bricks_check("odbc")
```

## 1. Where R looks for `.Renviron`

bricks reads its configuration (`DATABRICKS_HOST`, `DATABRICKS_WAREHOUSE_ID`,
`DATABRICKS_TOKEN`) from the environment, typically your user `.Renviron`. On
Windows, R's home directory (`~`) is often the **Documents** folder, which may
be redirected into OneDrive:

```{r}
path.expand("~")
#> "C:/Users/you/OneDrive - Org/Documents"   # not C:/Users/you !
```

R reads `~/.Renviron` at startup, so a `.Renviron` you saved in
`C:/Users/you/` will be ignored if `~` points somewhere else. Two fixes:

* **Recommended:** set `HOME` as a *Windows user environment variable* (not a
  line inside `.Renviron` -- that is too late, because R resolves `HOME` before
  reading the file):

  ```powershell
  [Environment]::SetEnvironmentVariable("HOME", "C:\Users\you", "User")
  ```

  Then fully restart your IDE so it inherits the new value, and move your
  `.Renviron` to `C:/Users/you/.Renviron`.

* **Or** simply put `.Renviron` wherever `path.expand("~")` reports.

## 2. A trailing newline matters

R silently drops the last line of `.Renviron` if the file does not end in a
newline. If `DATABRICKS_TOKEN` is your last line and the token appears unset,
this is the usual cause. Ensure a blank final line, or check:

```{r}
nzchar(Sys.getenv("DATABRICKS_TOKEN"))
#> TRUE
```

## 3. Some IDE kernels skip `.Renviron`

Some editor R kernels load `.Rprofile` at startup but not the user `.Renviron`.
If `bricks_check()` shows your variables unset even though the file is correct,
force the load from `.Rprofile` (which *is* read):

```{r}
# in ~/.Rprofile
if (file.exists("~/.Renviron")) readRenviron("~/.Renviron")
```

## 4. PAT vs. CLI OAuth

The odbc backend authenticates via `odbc::databricks()`, which accepts either a
`DATABRICKS_TOKEN` personal access token **or** the OAuth token from a matching
Databricks CLI profile. A `.databrickscfg` that only contains OAuth settings
(host / account_id / workspace_id / auth_type, no token) is *not* usable by the
odbc driver on its own -- generate a PAT (Databricks: User Settings ->
Developer -> Access tokens) and set `DATABRICKS_TOKEN`.

## 5. The Spark backends need local Spark

The `databricks` and `databricks_connect` backends run through `sparklyr`. On
Windows, local Spark additionally requires the **Microsoft Visual C++ 2010 SP1
x64 Redistributable**. If it is missing, `bricks_check("databricks")` flags it
and `bricks_connect("databricks")` aborts with a
`bricks_error_spark_unavailable` condition pointing at the fix:

1. Install `vcredist_x64.exe` from
   <https://www.microsoft.com/en-us/download/details.aspx?id=26999>.
2. Restart R.
3. `sparklyr::spark_install()`.

For plain SQL-warehouse queries you do not need Spark at all -- use the default
`odbc` backend.
